---
title: "Main"
author: "Group 4"
date: "2026-04-08"
output: html_document
runtime: shiny
---

Introduction

When we think about cars, we often focus on style, performance, or brand reputation but behind every vehicle is a combination of features that ultimately determine its price. For this project, we chose to analyze the Car Features and MSRP dataset from Kaggle, which contains detailed information about thousands of vehicles sold in the U.S. market. Each row represents a single car, and each column describes a specific characteristic such as horsepower, fuel efficiency, drivetrain type, number of doors, market category, and more.

This dataset is ideal for exploring how different design, performance, and mechanical features influence a car’s Manufacturer’s Suggested Retail Price (MSRP). Because it includes both numerical and categorical variables, it allows for a wide range of visualizations and comparisons. Our goal is to identify which features have the strongest relationship with MSRP and to understand how different types of vehicles cluster in the market.

To guide this analysis, we developed one main domain question and several sub‑questions that focus on specific features within the dataset.

Research Questions

Domain Question: What factors most strongly influence a car’s price(MSRP)? This question helps frame the overall purpose of the project identifying which vehicle characteristics are most closely associated with higher or lower prices. Sub Questions: What is the average price for two‑door cars as opposed to four‑door cars? This examines whether basic design features influence MSRP.

How do engine size, horsepower, or MPG relate to MSRP? This explores performance‑based predictors of price.

How does drivetrain type (FWD, RWD, AWD) influence average MSRP? This looks at how mechanical systems relate to cost.

Are there clusters or subgroups of cars (luxury vs. economy)? This helps identify whether market categories form distinct pricing groups.

Together, these questions allow for a complete analysis of how design, performance, and market identity contribute to vehicle pricing.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data <- read.csv("data.csv")
head(data)
```

```{r}
data

```

```{r}
#this is the highest priced vehicle for each group of Number.of.Doors

library(ggplot2)
ggplot(data, aes(Number.of.Doors, MSRP)) + 
  geom_col()
```


What factors most strongly influence a car’s price(MSRP)?
```{r}


library(shiny)
library(plotly)
library(dplyr)
ui <- fluidPage(
  titlePanel("Interactive MSRP Analysis"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "xvar",
        "Choose influencing factor:",
        choices = c("Engine Horsepower" = "Engine.HP",
                    "Highway MPG" = "highway.MPG",
                    "City MPG" = "city.mpg")
      ),
      
      selectInput(
        "make_filter",
        "Filter by Make:",
        choices = c("All", sort(unique(data$Make))),
        selected = "All"
      )
    ),
    
    mainPanel(
      plotlyOutput("msrp_plot")
    )
  )
)

server <- function(input, output) {
  
  filtered_data <- reactive({
    if (input$make_filter == "All") {
      data
    } else {
      data %>% filter(Make == input$make_filter)
    }
  })
  
  output$msrp_plot <- renderPlotly({
    df <- filtered_data()
    
    x_values <- df[[input$xvar]]
    
    plot_ly(
      data = df,
      x = ~x_values,
      y = ~MSRP,
      type = "scatter",
      mode = "markers",
      text = ~paste0(
        "Make: ", Make,
        "<br>Model: ", Model,
        "<br>", input$xvar, ": ", x_values,
        "<br>MSRP: $", MSRP
      ),
      hoverinfo = "text"
    ) %>%
      layout(
        title = paste(input$xvar, "vs MSRP"),
        xaxis = list(title = input$xvar),
        yaxis = list(title = "MSRP")
      )
  })
}

shinyApp(ui = ui, server = server)
```


What is the average price for tow door cars a s opposed to four door cars?
```{r}
library(dplyr)

data %>%
  group_by(Number.of.Doors) %>%
  summarize(avg_MSRP = mean(MSRP, na.rm = TRUE)) %>%
  ggplot(aes(x = Number.of.Doors, y = avg_MSRP)) +
  geom_col()
#this checks the average of each type of Number.of.Doors
```


How does drivetrain type (FWD, RWD, AWD) influence average MSRP?
```{r}

library(dplyr)
library(ggplot2)
data %>%
  group_by(Driven_Wheels) %>%
  summarize(avg_MSRP = mean(MSRP, na.rm = TRUE)) %>%
  ggplot(aes(x = Driven_Wheels, y = avg_MSRP, fill = Driven_Wheels)) +
  geom_col() +
  labs(title = "Average MSRP by Drivetrain Type",
       x = "Drivetrain",
       y = "Average MSRP")
```

Next, we compared average MSRP across drivetrain types between FWD, RWD, and AWD. The results show a clear pattern that AWD vehicles have the highest average MSRP, likely due to the added mechanical complexity and their common use in luxury or performance models. RWD vehicles fall in the middle, often associated with sports sedans and performance‑oriented cars. FWD vehicles have the lowest average MSRP, which makes sense because they are typically used in economy and compact cars. This indicates that mechanical systems such as drivetrain type meaningfully influence vehicle pricing.

How do engine size, horsepower, or MPG relate to MSRP?
```{r}

library(ggplot2)
# Horsepower vs MSRP
ggplot(data, aes(x = Engine.HP, y = MSRP)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(title = "Horsepower vs MSRP",
       x = "Engine Horsepower",
       y = "MSRP")
# City MPG vs MSRP
ggplot(data, aes(x = city.mpg, y = MSRP)) +
  geom_point(alpha = 0.5, color = "firebrick") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(title = "City MPG vs MSRP",
       x = "City MPG",
       y = "MSRP")
# Highway MPG vs MSRP
ggplot(data, aes(x = highway.MPG, y = MSRP)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  labs(title = "Highway MPG vs MSRP",
       x = "Highway MPG",
       y = "MSRP")
```

To understand how performance features influence price, we examined three variables: horsepower, city MPG, and highway MPG. Horsepower vs. MSRP: There is a strong positive relationship between horsepower and MSRP. Vehicles with higher horsepower tend to be significantly more expensive. This aligns with expectations, since performance‑oriented engines cost more to produce and are typically found in luxury or sports models. City MPG vs. MSRP: City MPG shows a negative relationship with MSRP. Cars with higher fuel efficiency tend to be cheaper, while low‑MPG vehicles (such as sports cars and large SUVs) have higher MSRPs. Highway MPG vs. MSRP:\
Highway MPG follows the same pattern as city MPG. More efficient vehicles cluster at lower price points, while less efficient vehicles are priced higher. Overall, these results show that performance features are strong predictors of price. High‑power, low‑efficiency vehicles cost more, while economical vehicles tend to be more affordable.

Are there clusters or subgroups of cars (luxury vs. economy)?
```{r}

library(dplyr)
library(ggplot2)

data$luxury_flag <- ifelse(
  grepl("Luxury|Performance|Exotic", data$Market.Category),
  "Luxury",
  "Non-Luxury"
)
ggplot(data, aes(x = luxury_flag, y = MSRP, fill = luxury_flag)) +
  geom_boxplot() +
  labs(title = "Luxury vs Non-Luxury MSRP Comparison",
       x = "Category",
       y = "MSRP") +
  scale_fill_manual(values = c("Luxury" = "gold", "Non-Luxury" = "gray"))

```

To identify broader market patterns, we created a variable that separates cars into “Luxury” and “Non‑Luxury” categories based on the Market.Category column. A boxplot comparing MSRP across these groups shows a dramatic separation. Luxury vehicles have significantly higher MSRPs, forming a distinct cluster with much higher price ranges. Non‑luxury vehicles cluster at lower price points, with far less variation. This confirms that market identity is one of the strongest predictors of MSRP. Luxury brands and performance‑oriented categories consistently command higher prices.
