---
title: "Exploring the US Car Market"
author: "Group 4"
date: "04/15/2026"
output: slidy_presentation
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
knitr::opts_chunk$set(echo = FALSE)
data <- read.csv("data.csv")
head(data)

```

```{r}
# Drop Market Category — too many NAs and not relevant to our analysis
#data <- data %>%
#  select(-Market.Category)
# We will talk about this



# Standardize drive wheel labels to abbreviations
data <- data %>%
  mutate(Driven_Wheels = gsub("all wheel drive", "AWD", Driven_Wheels),
         Driven_Wheels = gsub("rear wheel drive", "RWD", Driven_Wheels),
         Driven_Wheels = gsub("front wheel drive", "FWD", Driven_Wheels),
         Driven_Wheels = gsub("four wheel drive", "4WD", Driven_Wheels))

# Consolidate verbose fuel type labels into clean, consistent categories
data <- data %>%
  mutate(Engine.Fuel.Type = case_when(
    Engine.Fuel.Type == "premium unleaded (required)"                  ~ "premium unleaded",
    Engine.Fuel.Type == "premium unleaded (recommended)"               ~ "premium unleaded",
    Engine.Fuel.Type == "flex-fuel (unleaded/E85)"                     ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded recommended/E85)" ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded required/E85)"    ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (unleaded/natural gas)"             ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded/E85)"             ~ "Flex-Fuel",
    TRUE ~ Engine.Fuel.Type
  ))

# Filter out unwanted transmission types and fuel types
data <- data %>%
  filter(!Transmission.Type %in% c("UNKNOWN", "DIRECT_DRIVE")) %>%
  filter(!Engine.Fuel.Type %in% c("", "natural gas", "electric"))


# Remove rows with missing values in key numeric columns, as well as cars that cost over a million, as they skew the dataset.
data <- data %>%
  filter(!is.na(Number.of.Doors),
         !is.na(Engine.Cylinders),
          !is.na(Engine.HP)) %>%
  filter(MSRP <= 1000000)

# this is to take out one outlire that is not accurate 

data <- data %>%
  filter(highway.MPG <= 200)

```


## Introduction

Have you ever wondered why some cars cost significantly more than others? Whether it’s horsepower, fuel efficiency, drivetrain type, or even the number of doors, each feature plays a role in determining a vehicle’s price. For this project, we analyzed the Car Features and MSRP dataset to explore how different characteristics relate to a car’s MSRP. This dataset includes information such as engine performance, MPG, drivetrain, market category, and more. Our goal is to understand which features most strongly influence price and whether clear subgroups exist within the U.S. car market.

## Dataset

- **Dataset:** Car Features and MSRP  
- **Source:** Kaggle  
- **Data scraped from:** Edmunds and Twitter  

> Edmunds is a well-known automotive inventory and information resource —
> a reliable source for U.S. market vehicle data.

## Inside the Dataset

- Each row represents a single car model
- Each column represents a specific vehicle feature

Variables cover:

- Price (MSRP)
- Engine performance (horsepower, cylinders)
- Fuel efficiency (city and highway MPG)
- Vehicle design (size, number of doors)
- Drivetrain type (FWD, RWD, AWD)
- Market category (Luxury, Performance, etc.)

## Data Cleaning — Step 1: Standardizing Drivetrain Labels

```{r echo=TRUE, eval=FALSE}
data <- data %>%
  mutate(Driven_Wheels = gsub("all wheel drive", "AWD", Driven_Wheels),
         Driven_Wheels = gsub("rear wheel drive", "RWD", Driven_Wheels),
         Driven_Wheels = gsub("front wheel drive", "FWD", Driven_Wheels),
         Driven_Wheels = gsub("four wheel drive", "4WD", Driven_Wheels))
```

Standardized the Drivetrain label by their abbreviation
- Well known
- Looked better on the eye for graph

## Data Cleaning — Step 2: Fuel Type Consolidation

```{r echo=TRUE, eval=FALSE}
data <- data %>%
  mutate(Engine.Fuel.Type = case_when(
    Engine.Fuel.Type == "premium unleaded (required)"    ~ "premium unleaded",
    Engine.Fuel.Type == "premium unleaded (recommended)" ~ "premium unleaded",
    Engine.Fuel.Type %in% c("flex-fuel (unleaded/E85)",
                             "flex-fuel (premium unleaded recommended/E85)",
                             "flex-fuel (premium unleaded required/E85)") ~ "Flex-Fuel",
    TRUE ~ Engine.Fuel.Type
  ))
```

## Data Cleaning — Step 3: Filtering Rare & Unknown Types

```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(!Transmission.Type %in% c("UNKNOWN", "DIRECT_DRIVE")) %>%
  filter(!Engine.Fuel.Type %in% c("", "natural gas", "electric"))
```

Keeps the analysis focused on standard gasoline and flex-fuel vehicles with known transmissions.

## Data Cleaning — Step 4: Dropping Rows with Missing Values

```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(!is.na(Number.of.Doors),
         !is.na(Engine.Cylinders),
         !is.na(Engine.HP))
```

Rows missing core variables like horsepower or cylinders can't contribute meaningfully to analysis.

## Data Cleaning — Step 5: Removing Outliers

```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(MSRP <= 1000000) %>%
  filter(highway.MPG <= 200)
```

- **MSRP > $1M** — exotic vehicles that compress the price scale for all other cars  
- **Highway MPG > 200** — a data entry error that distorts MPG visualizations


## Questions

Domain Question:\
What factors most strongly influence a car’s price (MSRP)?

Sub‑Questions:

1.  What is the average price for two‑door vs. four‑door cars?

2.  How do engine size, horsepower, or MPG relate to MSRP?

3.  How does drivetrain type (FWD, RWD, AWD) influence average MSRP?

4.  Are there clusters or subgroups of cars (luxury vs. economy)?

## Unfiltered Data

Some text about this table and how interesting the results are!

```{r cars, echo = TRUE}
summary(cars)
```

## Key Variable

Engine.HP – Horsepower

city.mpg / highway.MPG – Fuel efficiency

Driven_Wheels – Drivetrain type

Number.of.Doors – Vehicle design

Market.Category – Luxury, performance, exotic, etc.

MSRP – Manufacturer’s Suggested Retail Price

## Slide with Plot

```{r pressure}
plot(pressure)
```

some text about how cool my plot is.



## Top 10 Best Value Car Brands Based on MSRP, City MPG, and Highway MPG

```{r}
top_unique_cars <- data %>%
  filter(!is.na(MSRP), !is.na(city.mpg), !is.na(highway.MPG)) %>%
  mutate(
    msrp_score = (max(MSRP) - MSRP) / (max(MSRP) - min(MSRP)),
    city_score = (city.mpg - min(city.mpg)) / (max(city.mpg) - min(city.mpg)),
    highway_score = (highway.MPG - min(highway.MPG)) / (max(highway.MPG) - min(highway.MPG)),
    total_score = msrp_score + city_score + highway_score
  ) %>%
  arrange(desc(total_score)) %>%
  group_by(Make) %>%        # group by brand
  slice(1) %>%              # take BEST car per brand
  ungroup() %>%
  arrange(desc(total_score)) %>%  # rank again across brands
  select(Make, Model, Year, MSRP, city.mpg, highway.MPG, total_score) %>%
  slice(1:10)

top_unique_cars

```
## The thing is, this is a little misleading because it doesn’t tell the whole story...

### When we filter by make (and not model), I’ll show you what happens with this pie chart on the next slide.

## Nah, we’re joking—we know better than to use those “#DrShoemakerExtraPointsPlease”

### Yes, we’re desperate enough to hold you at pie chart point 😭


## Toyota and Honda form a duopoly on the chart, accounting for a mind-boggling 70% of the top 10 (kia keep up lil bro)
```{r}
top_unique_models <- data %>%
  filter(!is.na(MSRP), !is.na(city.mpg), !is.na(highway.MPG)) %>%
  mutate(
    msrp_score = (max(MSRP) - MSRP) / (max(MSRP) - min(MSRP)),
    city_score = (city.mpg - min(city.mpg)) / (max(city.mpg) - min(city.mpg)),
    highway_score = (highway.MPG - min(highway.MPG)) / (max(highway.MPG) - min(highway.MPG)),
    total_score = msrp_score + city_score + highway_score
  ) %>%
  arrange(desc(total_score)) %>%
  group_by(Make, Model) %>%   # Key change
  slice(1) %>%                # best version of each model
  ungroup() %>%
  arrange(desc(total_score)) %>%
  slice(1:10) %>%
  select(Make, Model, Year, MSRP, city.mpg, highway.MPG, total_score)


brand_counts <- top_unique_models %>%
  count(Make) %>%
  mutate(
    percent = n / sum(n),
    percent_label = paste0(round(percent * 100), "%")
  )

# Order (Toyota + Honda first)
brand_counts$Make <- factor(
  brand_counts$Make,
  levels = c("Toyota", "Honda", "Volkswagen", "Kia")
)

ggplot(brand_counts, aes(x = 1, y = percent, fill = Make)) +
  geom_col(width = 0.18) +
  geom_text(
    aes(label = percent_label),
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  
  # 🔥 Vibrant colors
  scale_fill_manual(values = c(
    "Toyota" = "#FF3B3B",     # bright red
    "Honda" = "#FF6B3B",      # orange-red
    "Volkswagen" = "#7B61FF", # vibrant purple
    "Kia" = "#00B8B8"         # teal
  )) +
  
  labs(
    title = "Brand Share of Top 10 Value Cars",
    subtitle = "Toyota and Honda dominate the distribution",
    x = "",
    y = "",
    fill = "Make"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )
  
```


## 