---
title: "Exploring the US Car Market"
author: "Group 4"
date: "04/15/2026"
output: slidy_presentation
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
knitr::opts_chunk$set(echo = FALSE)
data <- read.csv("data.csv")
head(data)

```

```{r}
# Standardize drive wheel labels to abbreviations
data <- data %>%
  mutate(Driven_Wheels = gsub("all wheel drive", "AWD", Driven_Wheels),
         Driven_Wheels = gsub("rear wheel drive", "RWD", Driven_Wheels),
         Driven_Wheels = gsub("front wheel drive", "FWD", Driven_Wheels),
         Driven_Wheels = gsub("four wheel drive", "4WD", Driven_Wheels))

# Consolidate verbose fuel type labels into clean, consistent categories
data <- data %>%
  mutate(Engine.Fuel.Type = case_when(
    Engine.Fuel.Type == "premium unleaded (required)"                  ~ "premium unleaded",
    Engine.Fuel.Type == "premium unleaded (recommended)"               ~ "premium unleaded",
    Engine.Fuel.Type == "flex-fuel (unleaded/E85)"                     ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded recommended/E85)" ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded required/E85)"    ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (unleaded/natural gas)"             ~ "Flex-Fuel",
    Engine.Fuel.Type == "flex-fuel (premium unleaded/E85)"             ~ "Flex-Fuel",
    TRUE ~ Engine.Fuel.Type
  ))

# Filter out unwanted transmission types and fuel types
data <- data %>%
  filter(!Transmission.Type %in% c("UNKNOWN", "DIRECT_DRIVE")) %>%
  filter(!Engine.Fuel.Type %in% c("", "natural gas", "electric"))


# Remove rows with missing values in key numeric columns, as well as cars that cost over a million, as they skew the dataset.
data <- data %>%
  filter(!is.na(Number.of.Doors),
         !is.na(Engine.Cylinders),
          !is.na(Engine.HP)) %>%
  filter(MSRP <= 1000000)

# this is to take out one outlire that is not accurate 

data <- data %>%
  filter(highway.MPG <= 200)

```


## Introduction
Have you ever wondered why some cars cost significantly more than others?
Using the **Car Features and MSRP** dataset from 1990-2017, we set out to answer our central question:

> **What factors most strongly influence a car's price?**

## Questions
Domain Question:\

What factors most strongly influence a car's price (MSRP)?

Sub‑Questions:\

1.  What is the average price for two‑door vs. four‑door cars?

2.  How do engine size, horsepower, or MPG relate to MSRP?

3.  How does drivetrain type (FWD, RWD, AWD) influence average MSRP?

4.  Which car offers the best overall value when considering MSRP, fuel efficiency, and performance?

## Why this Question ?

- Cars are a significant purchase for most people to make
- Price is misunderstood
- Help a buyer understand what drives costs to make a better decision

## Libraries
```{r echo=TRUE, eval=FALSE}
library(dplyr)      # Data manipulation
library(ggplot2)    # Data visualization
library(scales)     # Formatting
```

## Dataset

- **Dataset:** Car Features and MSRP  
- **Data scraped from:** Edmunds and Twitter  

> Edmunds is a well-known automotive inventory and information resource — a reliable source for U.S. market vehicle data.

- **Sources**  
https://www.kaggle.com/datasets/CooperUnion/cardataset/data  
https://www.edmunds.com

## Inside the Dataset
```{r echo=FALSE}
head(data)
```

## Key Variables 

- Engine.HP – Horsepower

- city.mpg / highway.MPG – Fuel efficiency

- Driven_Wheels – Drivetrain type

- Number.of.Doors – Vehicle design

- Market.Category – Luxury, performance, exotic, etc.

- MSRP – Manufacturer’s Suggested Retail Price


## Data Cleaning — Step 1: Standardizing Drivetrain Labels
```{r echo=TRUE, eval=FALSE}
data <- data %>%
  mutate(Driven_Wheels = gsub("all wheel drive", "AWD", Driven_Wheels),
         Driven_Wheels = gsub("rear wheel drive", "RWD", Driven_Wheels),
         Driven_Wheels = gsub("front wheel drive", "FWD", Driven_Wheels),
         Driven_Wheels = gsub("four wheel drive", "4WD", Driven_Wheels))
```

Standardized the Drivetrain label by their abbreviation
- Well known
- Looked better on the eye for graph

## Data Cleaning — Step 2: Fuel Type Consolidation
```{r echo=TRUE, eval=FALSE}
data <- data %>%
  mutate(Engine.Fuel.Type = case_when(
    Engine.Fuel.Type == "premium unleaded (required)"    ~ "premium unleaded",
    Engine.Fuel.Type == "premium unleaded (recommended)" ~ "premium unleaded",
    Engine.Fuel.Type %in% c("flex-fuel (unleaded/E85)",
                             "flex-fuel (premium unleaded recommended/E85)",
                             "flex-fuel (premium unleaded required/E85)") ~ "Flex-Fuel",
    TRUE ~ Engine.Fuel.Type
  ))
```

- Multiple verbose variants of the same fuel type were being treated as separate categories


## Data Cleaning — Step 3: Filtering Rare & Unknown Types
```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(!Transmission.Type %in% c("UNKNOWN", "DIRECT_DRIVE")) %>%
  filter(!Engine.Fuel.Type %in% c("", "natural gas", "electric"))
```

- Transmission types labeled UNKNOWN provide no analytical value
- Natural gas and electric vehicles follow different pricing structures than standard vehicles


## Data Cleaning — Step 4: Dropping Rows with Missing Values
```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(!is.na(Number.of.Doors),
         !is.na(Engine.Cylinders),
         !is.na(Engine.HP))
```

- Rows missing horsepower, cylinders, or doors cannot contribute to price analysis


## Data Cleaning — Step 5: Removing Outliers
```{r echo=TRUE, eval=FALSE}
data <- data %>%
  filter(MSRP <= 1000000) %>%
  filter(highway.MPG <= 200)
```

- **MSRP > $1M** — exotic vehicles that compress the price scale for all other cars  
- **Highway MPG > 200** — a data entry error that distorts MPG visualizations


## Two doors VS Four doors
```{r}
data %>%
  group_by(Number.of.Doors) %>%
  summarize(avg_MSRP = mean(MSRP, na.rm = TRUE)) %>%
  ggplot(aes(x = Number.of.Doors, y = avg_MSRP)) +
  geom_col()
#this checks the average of each type of Number.of.Doors
```

- Two‑door cars tend to have higher average MSRPs, often due to sports/performance models car.

## Average MSRP by Drivetrain type 
```{r}
data %>%
  group_by(Driven_Wheels) %>%
  summarize(avg_MSRP = mean(MSRP, na.rm = TRUE)) %>%
  ggplot(aes(x = Driven_Wheels, y = avg_MSRP, fill = Driven_Wheels)) +
  geom_col() +
  labs(title = "Average MSRP by Drivetrain Type",
       x = "Drivetrain",
       y = "Average MSRP")
```

- AWD vehicles have the highest average MSRP

- FWD vehicles have the lowest

## engine size, horsepower, MPG relationship with MSRP
```{r}
# Horsepower vs MSRP
ggplot(data, aes(x = Engine.HP, y = MSRP)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "Horsepower vs MSRP",
       x = "Engine Horsepower",
       y = "MSRP")
```

- Higher horsepower is strongly associated with higher MSRP.

## City MPG vs MSRP
```{r}
ggplot(data, aes(x = city.mpg, y = MSRP)) +
  geom_point(alpha = 0.5, color = "firebrick") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "City MPG vs MSRP",
       x = "City MPG",
       y = "MSRP")
```

- More fuel‑efficient cars tend to be cheaper.

## Highway MPG vs MSRP
```{r}
ggplot(data, aes(x = highway.MPG, y = MSRP)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "Highway MPG vs MSRP",
       x = "Highway MPG",
       y = "MSRP")
```


## Top 10 Best Value Car Brands Based on MSRP, City MPG, and Highway MPG

```{r}
top_unique_cars <- data %>%
  filter(!is.na(MSRP), !is.na(city.mpg), !is.na(highway.MPG)) %>%
  mutate(
    msrp_score = (max(MSRP) - MSRP) / (max(MSRP) - min(MSRP)),
    city_score = (city.mpg - min(city.mpg)) / (max(city.mpg) - min(city.mpg)),
    highway_score = (highway.MPG - min(highway.MPG)) / (max(highway.MPG) - min(highway.MPG)),
    total_score = msrp_score + city_score + highway_score
  ) %>%
  arrange(desc(total_score)) %>%
  group_by(Make) %>%        # group by brand
  slice(1) %>%              # take BEST car per brand
  ungroup() %>%
  arrange(desc(total_score)) %>%  # rank again across brands
  select(Make, Model, Year, MSRP, city.mpg, highway.MPG, total_score) %>%
  slice(1:10)

top_unique_cars

```


- Highway MPG shows the same negative relationship with MSRP.

## The thing is, this is a little misleading because it doesn’t tell the whole story...

### When we filter by make (and not model), I’ll show you what happens with this pie chart on the next slide.

## Nah, we’re joking—we know better than to use those “#DrShoemakerExtraPointsPlease”

### Yes, we’re desperate enough to hold you at pie chart point 😭


## Toyota and Honda form a duopoly on the chart, accounting for a mind-boggling 70% of the top 10 (kia keep up lil bro)
```{r}
top_unique_models <- data %>%
  filter(!is.na(MSRP), !is.na(city.mpg), !is.na(highway.MPG)) %>%
  mutate(
    msrp_score = (max(MSRP) - MSRP) / (max(MSRP) - min(MSRP)),
    city_score = (city.mpg - min(city.mpg)) / (max(city.mpg) - min(city.mpg)),
    highway_score = (highway.MPG - min(highway.MPG)) / (max(highway.MPG) - min(highway.MPG)),
    total_score = msrp_score + city_score + highway_score
  ) %>%
  arrange(desc(total_score)) %>%
  group_by(Make, Model) %>%   # Key change
  slice(1) %>%                # best version of each model
  ungroup() %>%
  arrange(desc(total_score)) %>%
  slice(1:10) %>%
  select(Make, Model, Year, MSRP, city.mpg, highway.MPG, total_score)


brand_counts <- top_unique_models %>%
  count(Make) %>%
  mutate(
    percent = n / sum(n),
    percent_label = paste0(round(percent * 100), "%")
  )

# Order (Toyota + Honda first)
brand_counts$Make <- factor(
  brand_counts$Make,
  levels = c("Toyota", "Honda", "Volkswagen", "Kia")
)

ggplot(brand_counts, aes(x = 1, y = percent, fill = Make)) +
  geom_col(width = 0.18) +
  geom_text(
    aes(label = percent_label),
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  
  # 🔥 Vibrant colors
  scale_fill_manual(values = c(
    "Toyota" = "#FF3B3B",     # bright red
    "Honda" = "#FF6B3B",      # orange-red
    "Volkswagen" = "#7B61FF", # vibrant purple
    "Kia" = "#00B8B8"         # teal
  )) +
  
  labs(
    title = "Brand Share of Top 10 Value Cars",
    subtitle = "Toyota and Honda dominate the distribution",
    x = "",
    y = "",
    fill = "Make"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )
  
```


## Summary


- Price is strongly influenced by **horsepower and drivetrain type**  
- **Higher horsepower → higher MSRP**, while **higher MPG → lower MSRP**  
- **AWD vehicles** tend to be the most expensive, while **FWD are the cheapest**  
- When ranking value, **Toyota and Honda dominate**, making up **70% of the top 10**  
- Overall, the best value cars balance **low price, high fuel efficiency, and solid performance**  
