---
title: "Exploring the US Car Market"
author: "Group 4"
date: "04/15/2026"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
Have you ever wondered why some cars cost significantly more than others? Whether it’s horsepower, fuel efficiency, drivetrain type, or even the number of doors, each feature plays a role in determining a vehicle’s price. For this project, we analyzed the Car Features and MSRP dataset to explore how different characteristics relate to a car’s MSRP. This dataset includes information such as engine performance, MPG, drivetrain, market category, and more.
Our goal is to understand which features most strongly influence price and whether clear subgroups exist within the U.S. car market.

## Questions
Domain Question:  
What factors most strongly influence a car’s price (MSRP)?

Sub‑Questions:

1. What is the average price for two‑door vs. four‑door cars?

2. How do engine size, horsepower, or MPG relate to MSRP?

3. How does drivetrain type (FWD, RWD, AWD) influence average MSRP?

4. Are there clusters or subgroups of cars (luxury vs. economy)?

## Unfiltered Data

Some text about this table and how interesting the results are!

```{r cars, echo = TRUE}
summary(cars)
```

## Key Variable 

Engine.HP – Horsepower

city.mpg / highway.MPG – Fuel efficiency

Driven_Wheels – Drivetrain type

Number.of.Doors – Vehicle design

Market.Category – Luxury, performance, exotic, etc.

MSRP – Manufacturer’s Suggested Retail Price

## Slide with Plot

```{r pressure}
plot(pressure)
```

some text about how cool my plot is.

------------------------------------------------------------------------

Three dashes make a new slide without having to put a title, and the three-# titles don't make a new slide, only the two and one.

### see?

still on the same slide.
